Development of a 3D display

### 2016-08-20 paper.tex
old updates

% todo: references
As discussed before, there is no shortage of projects that are similar to mine, but only
display 2D images. The name "POV display", short for persistence of vision display, has become
commonplace. Another popular name is "Propeller Clock" --- Propeller because it rotates like one,
and clock because the round shape is ideal for displaying an analog clock.

As far as I know, however, only two people have built a working version of a three-dimensional POV
display so far (and have published it on the internet).\footnote{It turned out that there was a
third one: When I posted a picture of my project on the internet, a reddit user told me about the
display he built years before.} Neither of these is being produced on a grand scale, instead they
are --- like most 2D versions --- side projects of hobbyists or engineers.
% todo - link or reference in footnote

@@

Persistence of Vision refers to the optical illusion that results whenever the image we look
at quickly changes. In such situations, we don't immediately see the new image, instead we
continue seeing the old image for a short timespan after it has vanished. If wikipedia is to
be believed, this timespan is about $\frac{1}{25}$ of a second, but of course it is going to
be different for each pair of eyes.

@@

For controlling the LEDs, I have bought a Teensy 3.1 microcontroller development board. It has a
small footprint --- ideal for the fast rotation --- and still packs quite some power: The
processor runs at 96 MHz at a register width of 32 bit. There are 34 configurable I/O pins.

@@

\subsection{Data Transmission}

@@

Because \itwoc uses pullup resistors instead of a push-pull drive, it takes a little moment for
the logic level to return to its normal state. In order to ensure correct data transmission, it is
necessary to set a relatively slow clock speed. The two \itwoc interfaces of the Teensy only
support clock speeds of up to 2.4 MHz --- and because \itwoc requires some overhead data to send
out the slave's addresses, the actual data rate will be even lower.

@@

This means that I'd need at least 5 parallel \itwoc lines to control all the LEDs with 8 bit
colour depth. Because each Teensy only has two of them, I'd need three Teensy boards, which would
be possible but very cumbersome.

@@

% todo: Find out if this is necessary in the end
\newpage
\subsection{LED drivers}

@@

After quite some searching around, I decided to use the TLC5927 by Texas Instruments. It is
essentially a shift register with some extra features. Like every normal shift register, it has a
serial data input (SDI), a serial data output (SDO), and a clock pin (CLK). Additionally, there is
a latch pin (LE), which serves the purpose of only refreshing the state of the LEDs once all data
has been sent.

@@

In order for the whole device to rotate, I needed a motor. Luckily, I could salvage an appropriate
part from an old RC motor boat. It's a big and heavy brushed motor. The fact that it has a
built-in fan and was water-cooled in the boat leads me to think that its power output is much
higher than what I need.

@@

\newpage
\section{Circuit Diagram}

\begin{figure}[h]
\vspace{4mm}
\includegraphics[width=\textwidth]{./images/schematic.png}
\vspace{-8mm}
\caption{The finished circuit diagram}
\vspace{6mm}
\end{figure}

The circuit diagram --- as well as the PCB layout --- were created in Eagle. Its free version is
limited to two-layer boards of sizes up to \SI{10}{\centi\meter} by \SI{8}{\centi\meter}, but
that's more than enough for my purposes.

@@

\begin{figure}[h]
\vspace{4mm}
\includegraphics[width=\textwidth]{./images/board-layout.png}
\vspace{-6mm}
\caption{The PCB layout in Eagle. Red = upper copper layer, blue = lower copper layer,
green = vias, grey = silkscreen.}
\vspace{6mm}
\end{figure}

@@

The defining factor for the board layout was the pin layout of the LED drivers. The outputs are
all at the bottom of the chip, while the data interface is at the top. This led me to the layout I
have now: the LEDs are in a row at the bottom of the board where they can be directly connected to
the driver outputs on the upper copper layer. The control lines are is routed on the back side of
the board behind the chips, with vias going to the front side to connect to each chip. The data
lines go straight from one chip to the next one on the upper copper layer, with the last output
going back to the board's data interface on the back side. The resistors are placed on the right
side of their respective driver IC, jumping across the data traces.

@@

There's no shortage of firms who produce PCBs according to a custom layout. One of the cheapest
providers to ship small production runs within a reasonable timeframe and with good production
quality turned out to be OSHPark. They only produce in multiples of 3, so I ordered 12 instead of
10 (which I'd have done anyway to have some spare boards). I submitted my order on the 5th of
October 2014 and paid US\$ 47. Ten days later I received a message that the boards were finished,
and after another ten days of waiting they arrived at my doorstep.

@@

The following program has been compiled with the Arduino IDE 1.0.6 and sent to the Teensy by the
Teensyduino 1.20 plugin. However, the most part of the program has been written in Sublime Text 2
with the stino plugin. This combination has a few advantages over simply using the Arduino IDE to
write all code, namely there's automatic code completion, support for several cursors, and a
search-and-replace function with regex suport. I installed and used these programs on Mac OS 10.9
and later 10.10.

@@

In the main loop of the program, this variable is continuously compared to \texttt{sinceMagnet}.
If \texttt{sinceMagnet} is bigger, the LEDs are refreshed by calling \texttt{sendData}. Within
that function, \texttt{nextPixelMicros} is incremented by \texttt{microsPerPixel}, causing the
program to wait for the display to turn another $\frac{1}{100}$ of a full circle. After that,
it'll display the next pixel.

The implementation of the described behaviour can be found in the appendix, where the whole source
code is to be found. Specifically, what I've just talked about is contained in the
\texttt{timerUpdate} function, as well as in the main \texttt{loop}.

@@

To save the image on the microcontroller, I used a simple three-dimensional array containing $100
\cdot 10 \cdot 6$ bytes of data. The first index stands for the angle of a pixel in hundredths of
a circle, the second index corresponds to the height and the third one describes the position of
each byte on a row of LEDs. Because there are 48 LEDs per row, 6 bytes can be used to describe the
row's state without wasting any storage space. The only way to be more efficient in terms of
storage would be to compresss data, which would require significantly more computing power.

Because the PCBs are not right above each other, the data that is being sent out in one refresh
don't correspond to the pixels that are above each other in the image. In order avoid having to
account for the angular offsets of each board while creating an image, I made the program do that
work automatically.

@@

\bibitem{iictut}
\emph{\itwoc Tutorial}, article, retrieved 2016-05-19
\newline \url{http://www.best-microcontroller-projects.com/i2c-tutorial.html}

@@

\bibitem{intervaltimer}
\emph{IntervalTimer}, documentation page by Paul Stoffregen, retrieved 2016-05-19
\newline \url{https://www.pjrc.com/teensy/td\_timing\_IntervalTimer.html}

\bibitem{attachinterrupt}
\emph{attachInterrput()}, documentation page, retrieved 2016-05-19
\newline \url{https://www.arduino.cc/en/Reference/attachInterrupt}

\bibitem{kickstarter}
Comment by Paul Stoffregen, retrieved 2016-05-19
\newline
\url{https://www.kickstarter.com/projects/paulstoffregen/teensy-30-32-bit-arm-cortex-m4-usable-in-arduino-a/comments?cursor=1654857}

>>>>

% todo: references
As discussed before, there is no shortage of projects that are similar to mine, but only display
2D images \cite{povdisp, attiny, propclock}. The name "POV display", short for persistence of
vision display, has become commonplace. Another popular name is "Propeller Clock" --- Propeller
because it rotates like one, and clock because the round shape is ideal for displaying an analog
clock.

As far as I know, however, only two people have built a working version of a three-dimensional POV
display so far and have published it online \cite{volumen, hotchkiss}.\footnote{It turned out that
there was a third one - When I wrote a forum post about the display, somebody made a comment about
a similar display he built years ago: \url{https://www.reddit.com/comments/2q9sg6/-/cn4i4rk}}
Neither of these is being produced on a grand scale, instead they are --- like most 2D versions
--- side projects of hobbyists or engineers.

@@

Persistence of Vision refers to the optical illusion that results whenever the image we look
at quickly changes. In such situations, we don't immediately see the new image, instead we
continue seeing the old image for a short timespan after it has vanished. If wikipedia is to
be believed, this timespan is about $\frac{1}{25}$ of a second, but of course it is going to
be different for each pair of eyes \cite{povwiki}.

@@

For controlling the LEDs, I have bought a Teensy 3.1 microcontroller development board
\cite{teensy}. It has a small footprint --- ideal for the fast rotation --- and still packs quite
some power. The processor runs at 96 MHz at a register width of 32 bit, and there are 34
configurable I/O pins.

@@

\subsection{LED drivers}

@@

Because \itwoc uses pullup resistors instead of a push-pull drive, it takes a little moment for
the logic level to return to its normal state. In order to ensure correct data transmission, it is
necessary to set a relatively slow clock speed. Even with an optimised library \cite{iiclib}, the
two \itwoc interfaces of the Teensy are limited to 2.4 MHz if the clock speed it kept at
\SI{96}{\mega\hertz} --- and because \itwoc requires some overhead data to send out the slave's
addresses, the actual data rate will be even lower.

@@

This means that I'd need at least 5 parallel \itwoc lines to control all the LEDs with 8 bit
colour depth. Because each Teensy only has two of them, I'd need three Teensy boards, which would
be possible but very cumbersome. The SPI lines, however, can send a bit at every other clock
cycle, leading to a bit rate of \SI{48}{\mega\hertz}.

@@

% todo: Find out if this is necessary in the end
\newpage

@@

After quite some searching around, I decided to use the TLC5927 by Texas Instruments \cite{tlc,
tlcdigikey}. It is essentially a shift register with some extra features. Like every normal shift
register, it has a serial data input (SDI), a serial data output (SDO), and a clock pin (CLK).
Additionally, there is a latch pin (LE), which serves the purpose of only refreshing the state of
the LEDs once all data has been sent.

@@

In order for the whole device to rotate, I needed a motor. Luckily, I could salvage an appropriate
part from an old RC motor boat. It's a big and heavy brushed motor. The fact that it has a
built-in fan and was water-cooled in the boat leads me to believe that it can take a lot more
power than what I need for the display.

@@

% \newpage
\section{Circuit Diagram}

\begin{figure}[h]
\vspace{0mm}
\includegraphics[width=\textwidth]{./images/schematic.png}
\vspace{-9mm}
\caption{The finished circuit diagram}
\vspace{4mm}
\end{figure}

The circuit diagram --- as well as the PCB layout --- were created in Eagle \cite{eagle}. Its free
version is limited to two-layer boards of sizes up to \SI{10}{\centi\meter} by
\SI{8}{\centi\meter}, but that's more than enough for my purposes.

@@

@@

The defining factor for the board layout was the pin layout of the LED drivers. The outputs are
all at the bottom of the chip, while the data interface is at the top. This led me to the layout I
have now: the LEDs are in a row at the bottom of the board where they can be directly connected to
the driver outputs on the upper copper layer. The control lines are routed on the back side of
the board behind the chips, with vias going to the front side to connect to each chip. The data
lines go straight from one chip to the next one on the upper copper layer, with the last output
going back to the board's data interface on the back side. The resistors are placed on the right
side of their respective driver IC, jumping across the data traces.

\begin{figure}[h]
\vspace{4mm}
\includegraphics[width=\textwidth]{./images/board-layout.png}
\vspace{-6mm}
\caption{The PCB layout in Eagle. Red = upper copper layer, blue = lower copper layer,
green = vias, grey = silkscreen.}
\vspace{-3mm}
\end{figure}

@@

There's no shortage of firms who produce PCBs according to a custom layout. One of the cheapest
providers to ship small production runs within a reasonable timeframe and with good production
quality turned out to be OSHPark \cite{oshpark}. They only produce in multiples of 3, so I ordered
12 instead of 10 (which I'd have done anyway to have some spare boards). I submitted my order on
the 5th of October 2014 and paid US\$ 47. Ten days later I received a message that the boards were
finished, and after another ten days of waiting they arrived at my doorstep.

@@

The following program has been compiled with the Arduino IDE 1.0.6 \cite{arduino} and sent to the
Teensy by the Teensyduino 1.20 plugin \cite{teensyduino}. However, the most part of the program
has been written in Sublime Text 2 \cite{sublime} with the stino plugin \cite{stino}. This
combination has a few advantages over simply using the Arduino IDE to write all code, namely
there's automatic code completion, support for several cursors, and a search-and-replace function
with regex suport. I installed and used these programs on Mac OS 10.9 and later 10.10.

@@

In the main loop of the program, this variable is continuously compared to \texttt{sinceMagnet}.
If \texttt{sinceMagnet} is bigger, the LEDs are refreshed by calling \texttt{sendData}. Within
that function, \texttt{nextPixelMicros} is incremented by \texttt{microsPerPixel}, causing the
program to wait for the display to turn another $\frac{1}{100}$ of a full circle. After that, the
next pixel will be displayed.

% TODO put source code in appendix
The implementation of the described behaviour can be found in the appendix, where the whole source
code is included. Specifically, what I've just talked about is contained in the
\texttt{timerUpdate} function, as well as in the main \texttt{loop}.

@@

To save the image on the microcontroller, I used a simple three-dimensional array containing $100
\cdot 10 \cdot 6$ bytes of data. The first index stands for the angle of a pixel in hundredths of
a circle, the second index corresponds to the height and the third one describes the position of
each byte on a row of LEDs. Because there are 48 LEDs per row, 6 bytes can be used to describe the
row's state without wasting any storage space.

Because the PCBs are not right above each other, the data that is being sent out at the same time
don't correspond to the pixels that are above each other in the image. In order avoid having to
account for the angular offsets of each board while creating an image, I made the program do that
work automatically.

@@

@@

% \bibitem{intervaltimer}
% \emph{IntervalTimer}, documentation page by Paul Stoffregen, retrieved 2016-05-19
% \newline \url{https://www.pjrc.com/teensy/td\_timing\_IntervalTimer.html}

% \bibitem{attachinterrupt}
% \emph{attachInterrput()}, documentation page, retrieved 2016-05-19
% \newline \url{https://www.arduino.cc/en/Reference/attachInterrupt}

% \bibitem{kickstarter}
% Comment by Paul Stoffregen, retrieved 2016-05-19
% \url{https://www.kickstarter.com/projects/paulstoffregen/teensy-30-32-bit-arm-cortex-m4-usable-in-arduino-a/comments?cursor=1654857}
